Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-30 paper.tex
Document X numbers for SAIGE

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-CoV-2 ($n$=4,384,310), 
demonstrating the scalability and flexibility of VCF Zarr.
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative highly parallel benchmarks on AWS and experimentation with GPUs.

@@

>>>>

\textbf{Results:}
Zarr is a format for storing
multi-dimensional data that is widely used across the sciences,
and is ideally suited to massively parallel processing.
We present the VCF Zarr specification, an encoding of the
VCF data model using Zarr, along with fundamental software infrastructure
for efficient and reliable conversion at scale.
We show how this format is far more efficient than
standard VCF based approaches,
and competitive with specialised methods for
storing genotype data in terms of compression ratios
and single-threaded calculation performance.
We present case studies on subsets of three large 
human datasets (Genomics England: $n$=78,195;
Our Future Health: $n$=651,050; 
All of Us: $n$=245,394) along with whole genome
datasets for Norway Spruce ($n$=1,063) and 
SARS-CoV-2 ($n$=4,384,310).
We demonstrate the potential for VCF Zarr to enable a new generation
of high-performance and cost-effective applications
via illustrative examples using cloud computing and GPUs.

@@

% BCF: 4200.508 / 1223.767 = 3.43
% VCF 9727.995 / 1223.767 = 7.95
